---
title: "Architecture Overview"
author: "fmridataset Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Architecture Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

fmridataset separates *how data is stored* from *how analyses consume it*.
This document gives you a mental model of the three-layer stack so you can
predict how components interact, choose the right class for a given task, and
extend the package without modifying core code.

# Layer Diagram

```
  ┌──────────────────────────────────────────────────────┐
  │  User layer                                          │
  │  fmri_group        fmri_study_dataset                │
  │  (multi-study)     (multi-subject)                   │
  ├──────────────────────────────────────────────────────┤
  │  Dataset layer                                       │
  │  matrix_dataset    fmri_file_dataset   latent_dataset│
  │  (in-memory)       (NIfTI / HDF5)      (embedding)   │
  ├──────────────────────────────────────────────────────┤
  │  Backend layer                                       │
  │  matrix_backend  nifti_backend  h5_backend           │
  │  zarr_backend    study_backend                       │
  └──────────────────────────────────────────────────────┘
        ↑ all backends implement the same five-method contract
```

Each layer only knows about the layer directly beneath it.
Analysis code that calls `get_data_matrix()` or `data_chunks()` works
identically whether the data lives in RAM, on disk, or in a cloud-hosted
Zarr store.

# The Backend Contract

Every backend must implement five generic functions:

```{r backend-contract}
backend_open(backend)            # open file handles / allocate resources
backend_close(backend)           # release resources
backend_get_dims(backend)        # return list(spatial = c(x,y,z), time = n)
backend_get_data(backend,        # return matrix [time x voxels]
                 rows   = NULL,
                 cols   = NULL)
validate_backend(backend)        # stop() on contract violations
```

`rows` selects timepoints and `cols` selects voxels; both default to "all".
Backends may add caching, memory-mapping, or chunked I/O behind those five
calls without any dataset-layer changes.
For a full walkthrough of writing and registering a backend see the
`backend-development-basics` vignette.

# Dataset Classes

## `matrix_dataset`

Wraps an in-memory `[time x voxels]` matrix.
Use it for simulated data, preprocessed ROI time series, or any situation
where the full dataset fits comfortably in RAM.

```{r matrix-dataset-example}
ds <- matrix_dataset(
  datamat    = matrix(rnorm(100 * 500), nrow = 100, ncol = 500),
  TR         = 2.0,
  run_length = c(50, 50)
)
```

## `fmri_file_dataset`

Points at NIfTI or HDF5 files without loading them.
The underlying `nifti_backend` or `h5_backend` loads voxel blocks on demand,
so construction is near-instantaneous even for large scan collections.

```{r file-dataset-example}
ds <- fmri_file_dataset(
  scans      = c("run1.nii.gz", "run2.nii.gz"),
  mask       = "brain_mask.nii.gz",
  TR         = 2.0,
  run_length = c(200, 200)
)
```

## `latent_dataset`

Stores data in a lower-dimensional embedding space (e.g., ICA components,
PCA scores) rather than voxel space.
The interface is identical to the other dataset classes; only the column
semantics differ.

```{r latent-dataset-example}
ds <- latent_dataset(
  loadings   = component_matrix,   # voxels x components
  scores     = score_matrix,       # time   x components
  TR         = 2.0,
  run_length = c(100, 100)
)
```

## `fmri_study_dataset`

Aggregates multiple single-subject datasets under one object.
Subject-level data stays lazy; you iterate over subjects via `data_chunks()`
with `runwise = TRUE` or pull one subject at a time.

```{r study-dataset-example}
study <- fmri_study_dataset(
  datasets    = list(sub01_ds, sub02_ds, sub03_ds),
  subject_ids = c("sub-01", "sub-02", "sub-03")
)
```

# Temporal Structure

Every dataset carries a `sampling_frame` that models the acquisition
timeline.

```{r sampling-frame-example}
# Constructed automatically, or explicitly:
sf <- sampling_frame(
  blocklens = c(150, 150),   # run lengths in TRs
  TR        = 2.0
)

get_TR(sf)              # 2.0
get_run_lengths(sf)     # c(150, 150)
get_total_duration(sf)  # 600 seconds
```

The sampling frame handles all timepoint-to-second and TR-index conversions,
so the rest of the codebase never does raw arithmetic on timing.
An `event_table` can be attached to any dataset; onset times are validated
against the sampling frame at assignment.

Run lengths also drive the `runwise` chunking mode: when you request
`data_chunks(ds, runwise = TRUE)` the iterator yields one `[time x voxels]`
block per run, boundaries already aligned to the sampling frame.

# Extension Points

## Adding a New Backend

Implement the five contract functions for your new class, then register it:

```{r custom-backend-skeleton}
my_backend <- function(source, ...) {
  structure(list(source = source, ...), class = c("my_backend", "storage_backend"))
}

backend_open.my_backend         <- function(b) { ... }
backend_close.my_backend        <- function(b) { ... }
backend_get_dims.my_backend     <- function(b) { ... }
backend_get_data.my_backend     <- function(b, rows = NULL, cols = NULL) { ... }
validate_backend.my_backend     <- function(b) { ... }
```

Once those five methods exist, your backend can be used inside any
`fmri_file_dataset` by passing an instance via the `backend` argument, or
inside a custom dataset subclass.

## BIDS + HDF5 as a Concrete Example

`fmri_file_dataset` with `h5_backend` is the recommended pattern for
preprocessed BIDS derivatives stored in HDF5.
The `h5_backend` uses chunk-aware reads aligned to the HDF5 chunk lattice,
so iterating through a BIDS cohort with `data_chunks()` achieves near-optimal
I/O without any changes to analysis code.

# Object Zoo

| Class | Constructor | Purpose |
|---|---|---|
| `matrix_dataset` | `matrix_dataset()` | In-memory matrix, full random access |
| `fmri_file_dataset` | `fmri_file_dataset()` | Lazy NIfTI / HDF5 file access |
| `latent_dataset` | `latent_dataset()` | Embedding / component space data |
| `fmri_study_dataset` | `fmri_study_dataset()` | Multi-subject container |
| `sampling_frame` | `sampling_frame()` | Temporal structure for one session |
| `matrix_backend` | `matrix_backend()` | In-memory backend (internal) |
| `nifti_backend` | `nifti_backend()` | NIfTI file backend |
| `h5_backend` | `h5_backend()` | HDF5 backend with chunk-aware I/O |
| `zarr_backend` | `zarr_backend()` | Zarr backend for cloud-native arrays |
| `study_backend` | `study_backend()` | Multi-subject lazy backend |

# Related Vignettes

- `fmridataset-intro` — hands-on introduction to constructors and data access
- `backend-development-basics` — step-by-step guide to writing a new backend
- `study-level-analysis` — multi-subject iteration patterns with `fmri_study_dataset`
