---
title: "BIDS H5 Archive: Compressing a Study into a Single File"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{BIDS H5 Archive: Compressing a Study into a Single File}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

An fMRI study lives as thousands of files: NIfTI images, events.tsv files,
confound regressors, JSON sidecars. The BIDS H5 archive captures an entire
study in **one compressed HDF5 file** that is queryable by subject, task,
session, and run. You download one file and start analyzing immediately.

```{r load-package, eval = TRUE}
library(fmridataset)
```

# Which compression mode should you use?

The archive supports two modes. Pick the one that matches your analysis:

| Mode | What it stores | Reconstruction | Best for |
|:-----|:---------------|:---------------|:---------|
| `parcellated` | Cluster averages `[T, K]` | Parcel-level only | ROI analyses, connectivity |
| `latent` | Basis `[T, K]` + loadings `[V, K]` | Full voxel resolution | Searchlight, fine-grained spatial |

Both modes store events, confounds, censor vectors, and BIDS metadata
alongside the data. The standard `fmridataset` API works on either.

# How do you create an archive?

`compress_bids_study()` reads a BIDS directory and streams scans one at a
time into the archive. Only one NIfTI is held in memory at once.

## Parcellated mode

```{r write-parcellated}
library(bidser)

bids <- bids_project("/path/to/my_study")
atlas <- neuroim2::read_vol("schaefer_400.nii.gz")

study <- compress_bids_study(
  bids,
  file = "my_study.h5",
  mode = "parcellated",
  clusters = atlas
)
```

The result is a `bids_h5_study_dataset` you can use immediately.

## Latent mode

Latent mode uses `fmrilatent::encode()` to compress each scan into a
low-rank basis + loadings representation that can be reconstructed back
to voxel resolution.

```{r write-latent}
library(fmrilatent)

study <- compress_bids_study(
  bids,
  file = "my_study_latent.h5",
  mode = "latent",
  encoding = spec_time_dct(k = 30),
  mask = brain_mask
)
```

For multi-subject studies that share a spatial atlas, a **shared template**
stores the loadings once and only keeps per-scan coefficients, significantly
reducing file size:

```{r write-template}
tpl <- parcel_basis_template(parcellation, basis_spec = basis_slepian(k = 10))

study <- compress_bids_study(
  bids,
  file = "my_study_template.h5",
  mode = "latent",
  template = tpl,
  mask = brain_mask
)
```

# How do you open an archive?

One function, one file:

```{r read-archive}
study <- bids_h5_dataset("my_study.h5")
study
```

```
#> <bids_h5_study_dataset>
#>   File: my_study.h5
#>   Mode: parcellated (400 parcels)
#>   Subjects: 20 | Tasks: nback, rest | Sessions: pre, post
#>   Scans: 80 | Total timepoints: 24000 | TR: 2s
```

# Exploring the study

BIDS metadata is directly accessible:

```{r explore-metadata}
participants(study)
tasks(study)
sessions(study)
scan_manifest(study)
```

`scan_manifest()` returns a tibble with one row per scan:

```
#>   scan_name                       subject task  session run n_time
#>   sub-01_ses-pre_task-nback_run-01 01     nback pre     01  300
#>   sub-01_ses-pre_task-rest_run-01  01     rest  pre     01  200
#>   ...
```

# Subsetting by task, subject, or session

Use `subset_bids_h5()` to carve out the data you need. This returns a new
`bids_h5_study_dataset` backed by the same file:

```{r subset-study}
nback <- subset_bids_h5(study, task = "nback")
sub01 <- subset_bids_h5(study, subject = "01")
pre_nback <- subset_bids_h5(study, task = "nback", session = "pre")
```

Subsetting is cheap --- it selects scan backends from the shared H5
connection, no data is copied.

# Accessing data

The standard `fmridataset` API works on the result:

```{r access-data}
mat <- get_data_matrix(nback)
dim(mat)
```

```
#> [1] 6000  400
```

That is `[total_timepoints, K]` where K is the number of parcels (or
components in latent mode). Per-subject data:

```{r per-subject}
mat01 <- get_data_matrix(nback, subject_id = "01")
dim(mat01)
```

```
#> [1] 600 400
```

## Events

The event table combines all scans with `task`, `session`, `subject_id`,
and both the BIDS `run` label and an internal `run_id`:

```{r events}
head(nback$event_table)
```

```
#>   onset duration trial_type run run_id subject_id task  session
#>   0.0   2.0      face       01  1      01         nback pre
#>   4.0   2.0      house      01  1      01         nback pre
#>   ...
```

## Confounds

```{r confounds}
conf <- get_confounds(study, scan_name = "sub-01_ses-pre_task-nback_run-01")
head(conf)
```

## Chunked iteration

Memory-efficient processing via `data_chunks()`:

```{r chunks}
chunks <- data_chunks(nback, nchunks = 10)
while (!is.null(chunk <- iterators::nextElem(chunks))) {
  # Process chunk$data [T, K_chunk]
}
```

## Group operations

Convert to `fmri_group` for per-subject analyses:

```{r group}
group <- study_to_group(nback)
```

# Latent-mode extras

When working with a latent-mode archive, three additional accessors
are available:

```{r latent-extras}
info <- encoding_info(study)
info$encoding_family
info$n_components
info$has_shared_template

loadings <- get_loadings(study, scan_name = "sub-01_task-nback_run-01")
dim(loadings)

recon <- reconstruct_voxels(study,
  scan_name = "sub-01_task-nback_run-01",
  rows = 1:10,
  voxels = roi_indices
)
dim(recon)
```

`reconstruct_voxels()` computes `basis %*% t(loadings) + offset` on the
fly, so you only materialize the slice you need.

# Parcellation metadata

For parcellated archives, `parcellation_info()` gives you the cluster
mapping:

```{r parcellation}
pinfo <- parcellation_info(study)
pinfo$n_parcels
pinfo$cluster_ids
```

This returns `NULL` for latent-mode archives.

# Next steps

- `vignette("fmridataset-intro")` for the core dataset API
- `vignette("study-level-analysis")` for multi-subject workflows without BIDS
- `vignette("backend-development-basics")` if you want to write a custom backend
- `?compress_bids_study`, `?bids_h5_dataset`, `?subset_bids_h5` for full parameter docs
