---
title: "Implementing an array source"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.txt
vignette: >
  %\VignetteIndexEntry{Implementing an array source}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
- albers.css
- albers.js
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

Every assay in a frame is backed by an *array source*. A source is a
serializable description of a two-dimensional numerical array plus the S3
methods that read from it. The package ships in-memory, NIfTI, HDF5 (through
`fmristore`), and experimental Zarr sources. A storage package adds its own
by implementing a small set of generics. It then passes the same conformance
checks the built-in sources pass.

This vignette implements a complete source backed by an `.rds` file, in about
sixty lines, and validates it. The file format is deliberately trivial so the
protocol is the whole lesson.

```{r}
library(fmridataset)
```

## The protocol

An array source is an S3 object with class `c("<your_class>", "array_source")`
and methods for these generics:

| Generic | Returns |
|---|---|
| `source_shape()` | Two non-negative integers: observations, features |
| `source_dtype()` | One storage dtype such as `"float64"`, `"float32"`, `"int16"`, `"logical"` |
| `source_chunks()` | Two positive integers bounded by the shape: the natural read grid |
| `source_capabilities()` | Unique strings, including `"block_slice"` and `"serializable"` |
| `source_fingerprint()` | One non-empty string identifying this descriptor and revision |
| `source_open()` | A handle inheriting from `"array_source_handle"` |
| `source_read()` | An observation-by-feature matrix for integer selectors, in requested order |
| `source_read_native()` | A native spatial object, or an error if the source has none |
| `source_close()` | Releases the handle |

The package provides `source_descriptor()`, `validate_array_source()`,
`source_realization_cost()`, and `content_hash()` on top of these; you do not
implement them.

Two rules hold everything together.

**A descriptor is not a handle.** The source object holds only plain
serializable values: paths, shapes, dtypes, sizes, timestamps, tokens. It
holds no open connections, environments, external pointers, or closures.
`validate_array_source()` rejects a descriptor that contains any of them.
Runtime state belongs in the handle that `source_open()` returns. It never
becomes semantic state.

**Fingerprints are cheap and content hashes are explicit.** The fingerprint
is computed once, at construction, and stored in the descriptor. Its inputs
are the descriptor fields and any physical evidence you can observe without
reading values. It must never hash the array. Value identity is
`content_hash()`, which the package computes by streaming your
`source_read()`.

## An RDS-backed source

The constructor reads the file once to learn the shape, keeps the file's size
and modification time as revision evidence, computes the fingerprint, and
validates itself. `canonical_sha256()` is the package's canonical hash and is
the right tool for a fingerprint.

```{r constructor}
rds_source <- function(path, chunks = NULL) {
  path <- normalizePath(path, winslash = "/", mustWork = TRUE)
  values <- readRDS(path)
  if (!is.matrix(values) || !is.numeric(values)) {
    stop("rds_source() requires a numeric matrix.", call. = FALSE)
  }
  shape <- as.integer(dim(values))
  if (is.null(chunks)) chunks <- pmax(1L, shape)
  info <- file.info(path)
  out <- structure(
    list(
      path = path,
      shape = shape,
      dtype = "float64",
      chunks = pmin(as.integer(chunks), pmax(1L, shape)),
      size = as.numeric(info$size),
      mtime = as.numeric(info$mtime),
      schema_version = 1L
    ),
    class = c("rds_source", "array_source")
  )
  out$fingerprint <- canonical_sha256(list(
    type = "rds_source",
    schema_version = out$schema_version,
    path = out$path,
    shape = out$shape,
    dtype = out$dtype,
    chunks = out$chunks,
    size = out$size,
    mtime = out$mtime
  ))
  validate_array_source(out)
}
```

The descriptor accessors return stored fields.

```{r accessors}
source_shape.rds_source <- function(x, ...) x$shape
source_dtype.rds_source <- function(x, ...) x$dtype
source_chunks.rds_source <- function(x, ...) x$chunks
source_capabilities.rds_source <- function(x, ...) {
  c("row_slice", "column_slice", "block_slice", "serializable")
}
source_fingerprint.rds_source <- function(x, ...) x$fingerprint
```

Opening re-observes the file and compares it with the evidence captured at
construction. When they differ the descriptor is stale. That is not an I/O
failure: the file opened fine, it is simply not the file the descriptor was
made from. The package reserves `fmridataset_error_source_stale` for this
case and `fmridataset_error_backend_io` for genuine failures. Callers rely on
the distinction. Extension packages raise both as plain conditions with those
classes.

```{r open}
rds_assert_fresh <- function(x) {
  info <- file.info(x$path)
  actual <- list(size = as.numeric(info$size), mtime = as.numeric(info$mtime))
  expected <- list(size = x$size, mtime = x$mtime)
  if (!identical(actual, expected)) {
    stop(structure(
      list(
        message = paste("RDS file changed since the descriptor was built:", x$path),
        call = NULL,
        source = list(type = "rds_source", path = x$path),
        expected = expected,
        actual = actual
      ),
      class = c("fmridataset_error_source_stale", "fmridataset_error", "error", "condition")
    ))
  }
  invisible(TRUE)
}

source_open.rds_source <- function(x, ...) {
  rds_assert_fresh(x)
  structure(
    list(source = x, values = readRDS(x$path)),
    class = c("rds_source_handle", "array_source_handle")
  )
}
source_close.rds_source <- function(x, ...) invisible(TRUE)
```

Reads take integer selectors on both axes and honor the requested order,
and return a zero-extent matrix for an empty selection. The selection law
(`inst/architecture/ADR-010-selection-algebra.md`) is checked at the
`source_read()` generic before your method runs: positions arrive whole, in
bounds, and never repeated, so a method need not defend against a repeated
or out-of-range position, though refusing one again is harmless. Reading
through the handle is the real implementation; reading from the descriptor
opens, reads, and closes.

```{r read}
source_read.rds_source_handle <- function(x, observations = NULL, features = NULL, ...) {
  values <- x$values
  if (is.null(observations)) observations <- seq_len(nrow(values))
  if (is.null(features)) features <- seq_len(ncol(values))
  if (any(observations < 1L | observations > nrow(values)) ||
    any(features < 1L | features > ncol(values))) {
    stop("rds_source selector is out of bounds.", call. = FALSE)
  }
  values[observations, features, drop = FALSE]
}

source_read.rds_source <- function(x, observations = NULL, features = NULL, ...) {
  handle <- source_open(x)
  on.exit(source_close(handle), add = TRUE)
  source_read(handle, observations = observations, features = features, ...)
}

source_read_native.rds_source <- function(x, observations = NULL, ...) {
  stop("rds_source has no native spatial representation.", call. = FALSE)
}
```

That is the whole source. In a package these methods would be registered in
the `NAMESPACE` with `S3method()`; in a vignette, defining them at top level
is enough for dispatch.

## Try it

```{r use}
set.seed(4)
reference <- matrix(round(rnorm(30), 2), nrow = 6, ncol = 5)
path <- file.path(tempdir(), "toy.rds")
saveRDS(reference, path)

src <- rds_source(path, chunks = c(2L, 5L))
source_descriptor(src)
```

Reads honor order and shape.

```{r use-read}
source_read(src, observations = c(6L, 1L), features = c(5L, 2L))
reference[c(6, 1), c(5, 2)]
dim(source_read(src, observations = integer(), features = 1:2))
```

The source drops straight into a frame. `explain()` reports it by class, and
views over it are as lazy as any other.

```{r use-frame}
frame <- fmri_frame(
  assays = list(bold = src),
  observations = data.frame(.obs_id = sprintf("t%02d", 1:6)),
  space = index_space(5L, ids = paste0("f", 1:5))
)
explain(frame)$assays$bold[c("source_type", "chunks")]
collect_assay(frame[c("t02", "t01"), "f5"])
```

The cost estimate the budget checks use comes for free from the shape and
dtype.

```{r use-cost}
source_realization_cost(src, observations = 1:2)[c("storage_bytes", "estimated_peak_bytes")]
```

## Validate it

`validate_array_source()` checks the descriptor contract: two-dimensional
shape, supported dtype, chunk grid bounded by the shape, the required
capabilities, a non-empty fingerprint, and no runtime state.

```{r validate}
validate_array_source(src)
```

A descriptor that smuggles in a function, however well-formed otherwise,
fails.

```{r validate-fail, error = TRUE}
leaky <- src
leaky$loader <- function() readRDS(leaky$path)
validate_array_source(leaky)
```

The behavioral laws live in the package's conformance helper,
`tests/testthat/helper-frame-conformance.R`. It is a function that takes a
source and its reference matrix. A storage package should copy that helper
into its own tests and call it on every source type it ships. It checks, in
order:

- the descriptor is valid and its shape matches the reference;
- the descriptor survives `serialize()` and `unserialize()` with the same
  fingerprint, shape, dtype, and chunks;
- opening and closing a handle is safe and leaves the fingerprint untouched;
- reads match the reference for full, reversed, permuted, and empty
  selections, and are pure;
- out-of-range and repeated selections raise an error, and views composed
  over the source read what the equivalent flat positions read;
- `content_hash()` equals the content hash of a memory source over the
  reference, is accepted by `identity_descriptor()` as a content receipt, and
  does not change the fingerprint.

The last law is the one that matters most for interoperability, and the
cheapest to check by hand.

```{r conformance}
restored <- unserialize(serialize(src, NULL))
identical(source_fingerprint(restored), source_fingerprint(src))

handle <- source_open(src)
class(handle)
source_close(handle)

identical(content_hash(src), content_hash(memory_source(reference)))
```

`counting_source()` wraps any source and records what reaches it. That is
how the zero-I/O laws are certified: build a frame over the wrapper, take
views, and assert the read count is still zero.

```{r counting}
counted <- counting_source(src)
view <- source_view(counted, observations = 1:2)
source_counts(counted)$reads
invisible(source_read(view))
source_counts(counted)[c("reads", "values")]
```

Staleness behaves as designed. Rewriting the file does not change the
fingerprint stored in the descriptor. Every read now fails with the stale
class, and a fresh descriptor has a new fingerprint.

```{r stale}
Sys.sleep(1.1)
saveRDS(reference * 2, path)
class(tryCatch(source_read(src, 1L, 1L), error = identity))[1]
identical(source_fingerprint(rds_source(path)), source_fingerprint(src))
```

## Composition comes for free

The protocol is uniform, so the package's compositions accept the new source
without knowing anything about it. `row_sharded_source()` presents several
sources as one array with stable shard IDs and an inspectable manifest. That
is how multi-run and multi-subject data are bound.

```{r sharded}
saveRDS(reference[1:2, ], run_1 <- file.path(tempdir(), "run-1.rds"))
saveRDS(reference[3:6, ], run_2 <- file.path(tempdir(), "run-2.rds"))
sharded <- row_sharded_source(
  list(rds_source(run_1), rds_source(run_2)),
  shard_ids = c("run-1", "run-2")
)
shard_manifest(sharded)[, c(".shard_id", ".start", ".end")]
locate_source_rows(sharded, c(5L, 1L))[, c(".observation", ".shard_id", ".local_observation")]
identical(content_hash(sharded), content_hash(memory_source(reference)))
```

`as_delarr()` wraps any source as a lazy `delarr` array, so chunk-aware
numerical plans can pull from it through `source_read()` under the same
budget.

```{r delarr}
lazy <- as_delarr(sharded)
dim(lazy)
```

```{r cleanup, include = FALSE}
unlink(c(path, run_1, run_2))
```

## Checklist for a storage package

Keep the descriptor serializable and put every handle behind `source_open()`.
Compute the fingerprint at construction from descriptor fields and cheap
physical evidence, never from values. Raise
`fmridataset_error_source_stale` when the evidence no longer matches. Raise
`fmridataset_error_backend_io` when the backend fails. Return blocks in the
requested order, non-dropping, and refuse out-of-range selectors. Run the
conformance helper on every source type before release. The frame, view,
binding, planning, persistence, and hashing machinery will then work without
a single special case.
