---
title: "Feature spaces and maps"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.txt
vignette: >
  %\VignetteIndexEntry{Feature spaces and maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
- albers.css
- albers.js
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

The columns of an fMRI array are voxels, or vertices, or parcels, or latent
components. A frame's *feature space* says which. It names every column with
a stable ID and knows how to turn a row of values back into a spatial object.
This vignette builds the space types the package ships with and shows how
views restrict them. It explains what spatial identity means, then moves a
frame from one space to another through an explicit, lazy map.

```{r}
library(fmridataset)
library(Matrix)
```

## A masked volume

`volume_space()` describes a packed volume: a 3-D grid, a voxel-to-world
affine, and a *support*, the linear indices of the voxels that are in play.
Feature IDs are derived from the support, so they survive masking and
reordering unchanged.

```{r volume}
grid <- c(4L, 4L, 2L)
mask <- array(FALSE, grid)
mask[2:3, 2:3, ] <- TRUE

brain <- volume_space(dim = grid, affine = diag(4), support = mask, template = "toy")
n_features(brain)
feature_ids(brain)
head(feature_data(brain), 3)
```

The feature table gives each voxel its linear index and grid coordinates.
Predicates in `select_features()` can use any of these columns.

The frame used for the rest of the vignette has six observations over this
masked volume.

```{r frame}
set.seed(2)
n_obs <- 6L
signal <- matrix(
  round(rnorm(n_obs * n_features(brain)), 2),
  nrow = n_obs, ncol = n_features(brain)
)
frame <- fmri_frame(
  assays = list(bold = signal),
  observations = data.frame(
    .obs_id = sprintf("vol-%02d", seq_len(n_obs)),
    run_id = "run-1",
    TR = 2
  ),
  space = brain
)
frame
```

## Views restrict the space

A view's feature space is the base space restricted to the selected features,
in the selected order. Restriction is an operation on the space itself, not
on a copy of its table. The restricted space still knows its grid, its
affine, and where its voxels sit.

```{r restrict}
pair <- frame[, c("voxel-11", "voxel-6")]
space(pair)$support
feature_ids(pair)
same_space(space(pair), restrict_space(brain, c(4L, 1L)))$same
```

Restricting to nothing is legal and gives a space with zero features, which
is what an empty `select_features()` returns.

```{r restrict-empty}
n_features(restrict_space(brain, integer()))
```

A row of the frame can always be reconstructed into the native spatial object.
For a volume that is a `neuroim2::NeuroVol` with `NA` outside the support.

```{r spatial-map}
map <- spatial_map(frame, "vol-01")
dim(map)
sum(!is.na(map))
```

The same works from a view, where the unselected voxels are also `NA`.

```{r spatial-map-view}
sum(!is.na(spatial_map(pair, "vol-01")))
```

## Same space, not merely same size

`same_space()` reports exact semantic identity: the two spaces must agree on
class, digest, and ordered feature IDs. `compatible_space()` is an alias for
the same check, kept for the 1.0 transition. Neither word means a map exists.

An independently constructed space with the same grid, affine, support, and
template is the same space.

```{r same-space}
again <- volume_space(dim = grid, affine = diag(4), support = mask, template = "toy")
same_space(brain, again)$same
```

Change the affine and the voxels have the same IDs but sit somewhere else in
the world. The IDs still agree; the digest does not; the spaces are different.

```{r different-space}
rescaled <- volume_space(
  dim = grid, affine = diag(c(2, 2, 2, 1)), support = mask, template = "toy"
)
report <- same_space(brain, rescaled)
report[c("same", "same_class", "same_digest", "same_feature_ids")]
```

`assert_compatible_space()` turns that report into a structured error, and it
is what `bind_observations()` and `map_features()` call before doing anything
else.

```{r assert-space, error = TRUE}
assert_compatible_space(brain, rescaled)
```

## Parcels over a volume

A `parcel_space()` is linked to a parent space through a membership matrix.
It has one row per parent feature and one column per parcel, holding
non-negative weights. Here the two slices of the masked volume become two
parcels. The atlas identity is part of every parcel's feature ID.

```{r parcel}
slice <- feature_data(brain)$k
membership <- sparseMatrix(
  i = seq_len(n_features(brain)), j = slice, x = 1,
  dims = c(n_features(brain), 2L)
)
slabs <- parcel_space(
  parent = brain,
  parcel_ids = c("lower", "upper"),
  membership = membership,
  atlas = "slab",
  data = data.frame(id = c("lower", "upper"), label = c("Lower slab", "Upper slab"))
)
feature_ids(slabs)
parcel_aggregation(slabs)
```

The aggregation operator is derived from the membership: weighted means by
default, weighted sums on request. `parcel_space_from_atlas()` builds the same
object from a `neuroatlas` atlas aligned to a parent space.

## Moving a frame between spaces

A parcel space owns the canonical map from its parent, so `map_features()`
needs only the target. The result is a new frame in the parcel space. Its
assay is a *lazy* transformed source: nothing is aggregated until it is read.

```{r map-parcel}
by_parcel <- map_features(frame, target = slabs)
by_parcel
collect_assay(by_parcel)
```

Those are the per-slice means of `signal`, checked here against a direct
computation.

```{r map-parcel-check}
direct <- cbind(rowMeans(signal[, slice == 1]), rowMeans(signal[, slice == 2]))
all.equal(collect_assay(by_parcel), direct)
```

The transformation is recorded in the new frame's provenance, and a parcel
frame reconstructs back into the parent volume by spreading each parcel value
over its member voxels.

```{r map-parcel-provenance}
by_parcel$provenance
range(spatial_map(by_parcel, "vol-01"), na.rm = TRUE)
```

Any other linear transformation is an explicit `feature_map()`: a
target-by-source operator plus both spaces. The operator's dimensions are
checked against both. The source space must be the frame's space exactly.

```{r feature-map}
pairs <- index_space(2L, ids = c("pair-a", "pair-b"))
operator <- sparseMatrix(
  i = c(1, 1, 2, 2), j = c(1, 2, 3, 4), x = 0.5,
  dims = c(2L, n_features(brain))
)
pair_means <- feature_map(from = brain, to = pairs, operator = operator, map_type = "pair_mean")
pair_means
collect_assay(map_features(frame, map = pair_means))[1:2, ]
```

A map declared from a different space is refused even though the operator
would fit.

```{r feature-map-mismatch, error = TRUE}
wrong <- feature_map(from = rescaled, to = pairs, operator = operator)
map_features(frame, map = wrong)
```

## Bases and synthesis-only bases

A `basis_space()` is a representational axis over a parent space. It holds an
encoder that projects parent values onto components, a decoder that
synthesizes parent values from component scores, or both. When you hold only
a decoder, `basis_space_from_decoder()` derives the exact least-squares
encoder and validates that it is a left inverse.

```{r basis}
decoder <- matrix(0, n_features(brain), 3L)
decoder[1:4, 1L] <- 1
decoder[5:8, 2L] <- 1
decoder[, 3L] <- seq_len(n_features(brain)) / 8

components <- basis_space_from_decoder(
  parent = brain,
  component_ids = c("c1", "c2", "c3"),
  decoder = decoder,
  basis_type = "toy_basis"
)
basis_projection_info(components)[c("left_inverse_validated", "left_inverse_error")]
dim(basis_analysis(components))
```

With an encoder available, `map_features()` projects the frame onto the
components lazily, and `spatial_map()` synthesizes a volume back from scores.

```{r basis-map}
scores <- map_features(frame, target = components)
round(collect_assay(scores)[1:2, ], 3)
range(spatial_map(scores, "vol-01"), na.rm = TRUE)
```

Fitted dictionaries from ICA or dictionary learning are often rank-deficient,
so no exact left inverse exists. Ask for `encoder = "none"` and you get a
synthesis-only basis. It can reconstruct parent values from scores but has no
canonical map from the parent. A frame of scores must therefore be
constructed directly in that space.

```{r basis-synthesis-only}
synthesis_only <- basis_space_from_decoder(
  parent = brain,
  component_ids = c("c1", "c2", "c3"),
  decoder = decoder,
  encoder = "none"
)
is.null(basis_analysis(synthesis_only))
basis_projection_info(synthesis_only)$left_inverse_validated

score_frame <- fmri_frame(
  assays = list(scores = matrix(round(rnorm(n_obs * 3), 2), n_obs, 3L)),
  observations = observations(frame),
  space = synthesis_only
)
range(spatial_map(score_frame, "vol-01"), na.rm = TRUE)
```

Because the encoder is absent, `vectorize_space()` on this basis raises an
error that says so, and `map_features(frame, target = synthesis_only)` has no
operator to build a map from.

## Composite spaces

A `composite_space()` concatenates named child spaces into one feature axis,
for example left cortex, right cortex, and subcortical volume. Feature IDs
are part-qualified and the routing order is authoritative. Each child stays
the authority for its own identity and reconstruction.

```{r composite}
mixed_space <- composite_space(
  list(cortex = slabs, latent = components),
  composite_type = "toy_composite"
)
feature_ids(mixed_space)
feature_data(mixed_space)[, c(".feature_id", ".part", ".part_index")]
```

A frame over a composite space is selected and reconstructed like any other.
Selecting one part yields a composite space that holds only that part, and a
spatial map returns one native object per part.

```{r composite-frame}
mixed <- fmri_frame(
  assays = list(mixed = cbind(collect_assay(by_parcel), collect_assay(scores))),
  observations = observations(frame),
  space = mixed_space
)
latent_only <- select_features(mixed, .part == "latent")
feature_ids(latent_only)
composite_part_names(space(latent_only))

parts <- spatial_map(mixed, "vol-01")
names(parts$parts)
```

## Where the package stops

`fmridataset` owns spatial identity, restriction, the linear map descriptor,
and lazy application of that map. It does not fit bases, choose atlases, or
run models. `neuroatlas` supplies parcellations. `fmrilatent` supplies fitted
bases through `basis_space_from_fmrilatent()`. `neurosurf` supplies surface
geometry for `surface_space_from_neurosurf()`. Each hands the package a
serializable space; the package hands back frames whose columns still mean
what they meant.
