---
title: "Getting Started with fmridataset"
author: "fmridataset Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started with fmridataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  warning = FALSE,
  message = FALSE
)
library(fmridataset)
```

# Motivation

fMRI analyses involve heterogeneous sources: NIfTI files, BIDS datasets, preprocessed matrices, and HDF5 archives. fmridataset provides a unified interface that abstracts format differences so the same functions work regardless of how data is stored.

# Creating a Dataset

The simplest starting point is a matrix dataset, which wraps an in-memory matrix with temporal metadata.

```{r create-dataset}
set.seed(42)
mat <- matrix(rnorm(150 * 1000), nrow = 150, ncol = 1000)

ds <- matrix_dataset(
  datamat = mat,
  TR = 2.0,
  run_length = c(75, 75)
)

print(ds)
```

The dataset holds 150 timepoints split into two runs of 75, with a 2-second TR.

# Accessing Data

`get_data_matrix()` returns a standard timepoints-by-voxels matrix. You can retrieve all runs or a single run by index.

```{r get-data}
all_data <- get_data_matrix(ds)
cat("Full dimensions:", dim(all_data), "\n")

run1 <- get_data_matrix(ds, run_id = 1)
cat("Run 1 dimensions:", dim(run1), "\n")
```

The returned matrix is always in timepoints x voxels orientation.

# Temporal Structure

Every dataset contains a `sampling_frame` that records run boundaries, TR, and duration.

```{r sampling-frame}
sf <- ds$sampling_frame
cat("TR:", get_TR(sf), "seconds\n")
cat("Runs:", n_runs(sf), "\n")
cat("Run lengths:", get_run_lengths(sf), "timepoints\n")
cat("Total duration:", get_total_duration(sf), "seconds\n")

# First six acquisition times (seconds)
head(samples(sf))
```

`blockids()` maps each timepoint to its run index, which is useful for run-specific indexing.

# Event Tables

Experimental design attaches to the dataset as a data frame in `$event_table`.

```{r event-table}
events <- data.frame(
  onset     = c(10, 30, 50, 70, 110, 130, 150, 170),
  duration  = 2,
  trial_type = rep(c("faces", "houses"), 4),
  run       = c(1, 1, 1, 1, 2, 2, 2, 2)
)

ds$event_table <- events
head(ds$event_table)
```

Event onsets are in seconds and align with the sampling frame's time axis.

# Chunked Processing

For large datasets, `data_chunks()` partitions voxels into memory-manageable pieces without loading the entire matrix at once.

```{r chunked-processing}
chunks <- data_chunks(ds, nchunks = 4)

results <- lapply(chunks, function(chunk) {
  colMeans(chunk$data)
})

voxel_means <- do.call(c, results)
cat("Computed means for", length(voxel_means), "voxels\n")
```

Each element of `results` corresponds to one chunk; `do.call(c, ...)` reassembles them in voxel order.

# Run-wise Processing

Set `runwise = TRUE` to get one chunk per run. This is appropriate for analyses that must respect run boundaries such as detrending or temporal filtering.

```{r runwise-processing}
run_chunks <- data_chunks(ds, runwise = TRUE)

run_stats <- lapply(run_chunks, function(chunk) {
  cat("Run", chunk$chunk_num, ":", nrow(chunk$data), "timepoints\n")
  rowMeans(chunk$data)
})

cat("Processed", length(run_stats), "runs\n")
```

Each chunk's `$data` contains only that run's timepoints.

# File-based Datasets

When working with NIfTI files, use `fmri_file_dataset()`. Data remains on disk until explicitly accessed.

```{r file-dataset, eval=FALSE}
ds_files <- fmri_file_dataset(
  scans = c("/path/to/run1.nii.gz", "/path/to/run2.nii.gz"),
  mask  = "/path/to/mask.nii.gz",
  TR    = 2.0,
  run_length = c(180, 180)
)

# Inspect metadata without loading data
print(ds_files)

# Load one run only
run1 <- get_data_matrix(ds_files, run_id = 1)
```

Use `run_id` to load only the run you need, which keeps peak memory usage low.

# See Also

- `vignette("architecture-overview")` - Design principles and backend extensibility
- `vignette("h5-backend-usage")` - Efficient HDF5 storage for large datasets
- `vignette("study-level-analysis")` - Multi-subject studies and group analyses

# Session Information

```{r session-info}
sessionInfo()
```
