---
title: "Frames: aligned fMRI data with explicit identity"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.txt
vignette: >
  %\VignetteIndexEntry{Frames: aligned fMRI data with explicit identity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
- albers.css
- albers.js
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

An fMRI analysis moves the same numbers through many hands: a preprocessing
pipeline, a mask, a model, a parcellation, a file on disk. At each step the
rows and columns of the array still mean something, and that meaning is easy
to lose. `fmridataset` keeps it. Its one data container, the `fmri_frame`,
holds an observation-by-feature array together with what each observation and
each feature *is*, and refuses operations that would silently break that link.

This vignette builds a frame from a small synthetic matrix and walks through
the everyday operations: selecting by ID and by metadata, reading values
lazily, inspecting a frame without reading it, describing its run structure,
and binding frames together. It ends with the short list of promises every
frame keeps.

```{r}
library(fmridataset)
```

## Build a frame

A frame needs three things: at least one assay (the numbers), observation
metadata with a stable ID per row, and a feature space that says what the
columns are. Here the columns are the voxels of a tiny 3 by 3 by 2 volume,
and the rows are eight volumes acquired in two runs.

```{r build}
set.seed(1)
voxels <- volume_space(dim = c(3, 3, 2), affine = diag(4), template = "toy")
n_obs <- 8L
signal <- matrix(
  round(rnorm(n_obs * n_features(voxels)), 2),
  nrow = n_obs, ncol = n_features(voxels)
)

frame <- fmri_frame(
  assays = list(bold = signal),
  observations = data.frame(
    .obs_id = sprintf("vol-%02d", seq_len(n_obs)),
    run_id = rep(c("run-1", "run-2"), each = 4),
    TR = 2,
    motion = round(runif(n_obs), 2)
  ),
  space = voxels
)
frame
```

The `.obs_id` column supplies the observation IDs. The volume space supplies
the feature IDs, one per voxel in its support, and derives a feature table
from them.

```{r metadata}
observations(frame)
head(features(frame), 3)
```

Nothing about the matrix itself was trusted for identity. If `signal` had
carried row names that disagreed with `.obs_id`, construction would have
failed rather than picking one.

## Select by ID or by metadata

Square-bracket selection takes IDs on either axis and returns a *view*: a
lightweight object that remembers which rows and columns of the parent frame
it stands for. The requested order is kept on every axis.

```{r select-id}
view <- frame[c("vol-08", "vol-01"), c("voxel-5", "voxel-2")]
view
observation_ids(view)
feature_ids(view)
collect_assay(view)
```

Metadata predicates do the same job when you know a property rather than an
ID. `filter_obs()` evaluates an expression against the observation table and
`select_features()` against the feature table.

```{r select-metadata}
run_2 <- filter_obs(frame, run_id == "run-2")
observation_ids(run_2)

still <- filter_obs(frame, motion < 0.5)
observation_ids(still)

top_slice <- select_features(frame, k == 2)
feature_ids(top_slice)
```

Views compose. Selecting from a view produces another view over the same base
frame, and the feature space follows along: the space of `top_slice` is the
same volume restricted to the nine voxels of its upper slice.

```{r select-compose}
corner <- top_slice[, c("voxel-18", "voxel-10")]
space(corner)$support
```

Selectors that could mean two things are rejected. Duplicated IDs, unknown
IDs, and mixed positive and negative positions all raise a structured error
instead of a guess.

```{r select-errors, error = TRUE}
frame[c("vol-01", "vol-01"), ]
frame[, "voxel-99"]
```

## Views are lazy

A view stores a selection, not values. To see that nothing is read until you
ask, wrap the matrix in a `counting_source()`, the package's read
instrumentation for conformance tests, and build a frame over it.

```{r lazy}
counted <- counting_source(signal)
lazy <- fmri_frame(
  assays = list(bold = counted),
  observations = observations(frame),
  space = voxels
)

subset <- filter_obs(lazy, run_id == "run-1")[, c("voxel-1", "voxel-2")]
source_counts(counted)$reads

values <- collect_assay(subset)
dim(values)
source_counts(counted)[c("reads", "values")]
```

Filtering and subsetting issued no reads. `collect_assay()` issued one, and
it read exactly the eight values the view stands for, not the full matrix.
The same holds for file-backed sources, where the difference is a small
selection against a large volume on disk.

`collect_assay()` also checks an estimated memory cost against a budget before
it reads anything, so a request that would not fit fails early and cheaply.

```{r budget, error = TRUE}
collect_assay(frame, memory_budget = 10)
```

## Inspect without reading

`explain()` returns a bounded, serializable summary of a frame or view. It
reports the shape, the identities of the axes and space, the sources behind
each assay, and what realizing them would cost. It never reads assay values.

```{r explain}
report <- explain(frame)
report$shape
report$space
report$assays$bold[c("source_type", "dtype", "chunks", "realization_bytes")]
```

The digests are typed. The schema digest covers the column and space
contracts, the semantic digest covers the whole source-free manifest, and the
assay fingerprint identifies the physical source. None of them hashes the
numbers; `vignette("persistence-and-import")` explains when you want one that
does.

```{r explain-digests}
substr(unlist(report$digests), 1, 12)
```

## Run structure is metadata

A frame carries no acquisition timing of its own. When its observations are
volumes acquired in runs, that is a fact recorded in the observation table,
under a small validated contract: a run column, an optional `TR` in seconds
that is constant within each run, and an optional logical `censor` column.
`temporal_schema()` derives the description from those columns each time it is
asked, so it can never go stale.

```{r temporal}
has_temporal_schema(frame)
schema <- temporal_schema(frame)
schema
schema$run_lengths
```

The run column is discovered rather than assumed: a relation to a run entity
wins, then `scan_id`, then `run_id`, and an explicit `run_col` overrides all
three. When the `fmrihrf` package is installed, `as_sampling_frame()` builds the
run-length encoded `sampling_frame` that design and modelling code consumes.

```{r sampling-frame, eval = requireNamespace("fmrihrf", quietly = TRUE)}
as_sampling_frame(frame)
```

Because the schema follows the observation order, a filtered view still
describes itself correctly.

```{r temporal-view}
temporal_schema(still)
```

A view whose runs are interleaved is a legal frame, but a sampling frame is a
run-length encoding and cannot represent it. The conversion refuses rather
than reordering behind your back.

```{r temporal-interleaved, error = TRUE}
interleaved <- frame[c("vol-01", "vol-05", "vol-02"), ]
temporal_schema(interleaved)$contiguous
as_sampling_frame(interleaved)
```

Frames without run structure, such as a matrix of beta estimates, are
first-class; for them `has_temporal_schema()` is simply `FALSE`.

## Durable and ephemeral IDs

Every axis has a declared ID policy. The default, `require`, means the caller
supplied the IDs, as `.obs_id` did above. Both axes of `frame` are durable.

```{r durable}
ids_are_durable(frame)
axis_id_policy(observation_axis(frame))$policy
```

Importers that mint IDs from declared keys use the `deterministic` policy,
which derives a SHA-256 ID from a namespace and the typed key values. The same
keys always produce the same IDs, so two independent imports of the same data
agree.

```{r deterministic}
scans <- axis_frame(
  data.frame(subject = "01", run = rep(1:2, each = 4), volume = rep(1:4, 2)),
  axis = "observation",
  id_policy = "deterministic",
  id_namespace = "study:v1",
  id_keys = c("subject", "run", "volume")
)
substr(axis_ids(scans)[1:2], 1, 24)
axis_id_policy(scans)[c("policy", "namespace", "keys")]
```

For quick exploratory work you can ask for `ephemeral` IDs. They are UUIDs
with a visible `ephemeral-` prefix, and a frame that carries them cannot be
persisted or given a certified semantic identity until a durable axis
replaces them. Omitting `space` gives the feature axis the same treatment.

```{r ephemeral, error = TRUE}
scratch <- fmri_frame(
  assays = list(bold = signal),
  observations = axis_frame(data.frame(x = seq_len(n_obs)), id_policy = "ephemeral")
)
substr(observation_ids(scratch)[1], 1, 14)
ids_are_durable(scratch)
fds_frame_manifest(scratch)
```

## Bind frames along observations

`bind_observations()` stacks frames that share a feature space. The result is
lazy: assay values stay in their original sources and are routed by row when
read. Observation order follows the operand order, IDs must not collide, and
the spaces must be the same space, not merely the same size.

```{r bind}
run_1 <- filter_obs(frame, run_id == "run-1")
both <- bind_observations(run_2, run_1)
both
observation_ids(both)
temporal_schema(both)$run_lengths
```

Frame-level metadata is reconciled by an explicit policy. The default demands
equality; `"merge"` combines records recursively and fails only on genuine
conflict.

```{r bind-metadata, error = TRUE}
site_a <- fmri_frame(
  assays = list(bold = signal[1:4, ]),
  observations = observations(frame)[1:4, ],
  space = voxels,
  metadata = list(site = "A", scanner = list(vendor = "X"))
)
site_b <- fmri_frame(
  assays = list(bold = signal[5:8, ]),
  observations = observations(frame)[5:8, ],
  space = voxels,
  metadata = list(site = "A", scanner = list(field = "3T"))
)

bind_observations(site_a, site_b)

merged <- bind_observations(site_a, site_b, metadata_policy = "merge")
str(unclass(merged$metadata$scanner))
```

Every bind appends a provenance record naming its inputs, so a bound frame can
say where its rows came from.

```{r bind-provenance}
merged$provenance
```

A frame over a different volume, even one with identical dimensions, is
refused.

```{r bind-space, error = TRUE}
other_volume <- volume_space(dim = c(3, 3, 2), affine = diag(4), template = "other")
elsewhere <- fmri_frame(
  assays = list(bold = signal[1:4, ]),
  observations = observations(frame)[1:4, ],
  space = other_volume
)
bind_observations(site_a, elsewhere)
```

## What a frame promises

These are the contracts from the package's canonical data model
(`inst/architecture/ADR-001-canonical-data-model.md`), stated as the behavior
you can rely on.

- **Two axes, always.** An assay is observations by features. Factors and
  continuous variables are annotations on an axis, never a third dimension.
- **Every assay shares the axes.** All assays in a frame have the same
  observation IDs and the same feature IDs, in the same order.
- **IDs are stable and explicit.** Subsetting, reordering, and binding carry
  IDs through unchanged. Nothing regenerates them, and nothing aligns by
  position.
- **Spatial identity is semantic.** Two spaces are the same when their class,
  digest, and ordered feature IDs agree. Matching dimensions are not evidence.
- **Ambiguity fails early.** Duplicated selectors, conflicting dimnames,
  disagreeing metadata, and colliding IDs raise structured errors rather than
  being resolved by a rule you did not choose.
- **Execution is lazy and bounded.** Views and binds read nothing.
  Realization goes through an explicit budget, and `explain()` tells you the
  cost first.
- **Descriptors, not handles.** A frame contains serializable descriptors of
  its sources, never open files, environments, or closures, so it can be
  serialized and rebuilt elsewhere.

`vignette("feature-spaces")` covers the feature-space types and how to move a
frame between them. `vignette("persistence-and-import")` covers the FDS
manifest, HDF5 round trips, identity, and BIDS import.
