---
title: "Using the H5 Backend for Efficient fMRI Data Storage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the H5 Backend for Efficient fMRI Data Storage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to FALSE since we don't have actual H5 files in the package
)
```

# H5 Backend for fMRI Data

The `fmridataset` package supports efficient storage and access of fMRI data using HDF5 files through integration with the `fmristore` package. Each scan is stored as an H5 file that loads to an `H5NeuroVec` object, providing memory-efficient access to large neuroimaging datasets.

## Prerequisites

To use the H5 backend, you need to install the `fmristore` package:

```{r, eval=FALSE}
# Install from GitHub
devtools::install_github("bbuchsbaum/fmristore")
```

You also need the following dependencies:
- `neuroim2` for neuroimaging data structures
- `hdf5r` for HDF5 file operations

## Basic Usage

### Creating an fMRI Dataset from H5 Files

```{r}
library(fmridataset)

# Create an fMRI dataset from H5 files
dset <- fmri_h5_dataset(
  h5_files = c("scan1.h5", "scan2.h5", "scan3.h5"),
  mask_source = "mask.h5",  # Can also be a regular .nii file
  TR = 2.0,
  run_length = c(150, 150, 150)
)

# The dataset behaves like any other fmri_dataset
print(dset)
```

### Using a Pre-created H5 Backend

For more control, you can create an H5 backend directly:

```{r}
# Create H5 backend
backend <- h5_backend(
  source = c("functional_run1.h5", "functional_run2.h5"),
  mask_source = "brain_mask.h5",
  data_dataset = "data/elements",  # HDF5 dataset path for data
  mask_dataset = "data/elements",  # HDF5 dataset path for mask
  preload = FALSE  # Load data on-demand rather than eagerly
)

# Create dataset using the backend
dset <- fmri_dataset(
  scans = backend,
  TR = 2.5,
  run_length = c(200, 200),
  event_table = my_events
)
```

## H5 File Structure

The H5 backend expects HDF5 files with the following structure (compatible with `fmristore`):

```
/
├── space/
│   ├── dim          # Spatial dimensions [x, y, z, t]
│   ├── origin       # Volume origin coordinates
│   └── trans        # Transformation matrix
└── data/
    └── elements     # 4D fMRI data array
```

For mask files:
```
/
├── space/
│   ├── dim          # Spatial dimensions [x, y, z]
│   ├── origin       # Volume origin coordinates
│   └── trans        # Transformation matrix
└── data/
    └── elements     # 3D mask array
```

## Advantages of H5 Backend

1. **Memory Efficiency**: Data is loaded on-demand, reducing memory usage for large datasets
2. **Fast Access**: HDF5 provides efficient random access to data subsets
3. **Compression**: Built-in compression reduces storage requirements
4. **Cross-platform**: HDF5 files work across different operating systems
5. **Metadata Storage**: Rich metadata can be stored alongside the data

## Working with Large Datasets

The H5 backend is particularly useful for large datasets:

```{r}
# For very large datasets, use preload=FALSE (default)
large_backend <- h5_backend(
  source = paste0("scan_", 1:20, ".h5"),  # 20 scans
  mask_source = "mask.h5",
  preload = FALSE  # Data loaded only when accessed
)

# Create dataset
large_dset <- fmri_dataset(
  scans = large_backend,
  TR = 2.0,
  run_length = rep(300, 20)  # 20 runs of 300 timepoints each
)

# Data is loaded on-demand when you access it
data_matrix <- get_data_matrix(large_dset)
```

## Converting Existing Data to H5 Format

If you have existing NIfTI data, you can convert it to H5 format using `fmristore`:

```{r}
# This would be done outside of fmridataset, using fmristore directly
library(fmristore)
library(neuroim2)

# Load NIfTI data
nvec <- read_vec("functional.nii")

# Convert to H5 format
h5_file <- as_h5(nvec, file = "functional.h5", data_type = "FLOAT", compression = 6)

# Now you can use it with the H5 backend
dset <- fmri_h5_dataset("functional.h5", "mask.nii", TR = 2.0, run_length = 300)
```

## Performance Considerations

- Use `preload = TRUE` for small datasets that fit in memory
- Use `preload = FALSE` (default) for large datasets
- H5 files benefit from compression - use compression levels 4-6 for good balance of speed/size
- For optimal performance, ensure H5 files are stored on fast storage (SSD)

## Error Handling

The H5 backend provides informative error messages:

```{r, error=TRUE}
# File not found
try(h5_backend("nonexistent.h5", "mask.h5"))

# Invalid object types
try(h5_backend(list("not_an_h5neurovec"), "mask.h5"))

# Missing fmristore package
# (Error shown if fmristore is not installed)
```

## Integration with Existing Workflows

The H5 backend integrates seamlessly with existing `fmridataset` workflows:

```{r}
# All standard operations work the same way
mask <- get_mask(dset)
data <- get_data_matrix(dset)
sampling_frame <- get_sampling_frame(dset)

# Data chunking works efficiently with H5 backend
chunks <- data_chunks(dset, nchunks = 4)
```

The H5 backend provides an efficient way to work with large fMRI datasets while maintaining the familiar `fmridataset` interface. 