---
title: "Persistence, identity, and import"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.txt
vignette: >
  %\VignetteIndexEntry{Persistence, identity, and import}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
- albers.css
- albers.js
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
has_fmristore <- requireNamespace("fmristore", quietly = TRUE) &&
  "write_frame_h5" %in% getNamespaceExports("fmristore")
has_bidser <- requireNamespace("bidser", quietly = TRUE) &&
  utils::packageVersion("bidser") >= "0.5.0"
```

A frame that only lives in one R session is of limited use. This vignette
covers the three ways a frame meets the outside world: its backend-neutral
manifest, the HDF5 round trip that stores it, and the BIDS on-ramp that
creates one from an fMRIPrep derivative tree. Running through all three is the
question of identity: what it means for two frames, two spaces, or two arrays
to be *the same*.

```{r}
library(fmridataset)
```

The running example is a small volume-backed frame with two runs.

```{r frame}
set.seed(3)
voxels <- volume_space(dim = c(3, 3, 2), affine = diag(4), template = "toy")
n_obs <- 6L
signal <- matrix(round(rnorm(n_obs * 18), 2), n_obs, 18)
observations <- data.frame(
  .obs_id = sprintf("vol-%02d", seq_len(n_obs)),
  run_id = rep(c("run-1", "run-2"), each = 3),
  TR = 2
)
frame <- fmri_frame(
  assays = list(bold = signal),
  observations = observations,
  space = voxels,
  metadata = list(study = "toy")
)
```

## The FDS manifest

`fds_frame_manifest()` produces the FDS version 1 manifest: everything a
frame means, and nothing about where its numbers are stored. It holds the
axis IDs and tables, the feature space, the assay declarations with their
shapes and dtypes, entities, relations, typed tables, metadata, and
provenance. It performs no I/O and reads no assay values.

```{r manifest}
manifest <- fds_frame_manifest(frame)
names(manifest)
manifest$schema$id
manifest$assays$bold[c("name", "array", "dtype", "shape")]
```

`fds_manifest_digest()` canonicalizes the manifest and hashes it. This is the
frame's *semantic* identity. Because physical sources are not in the
manifest, the same frame has the same digest however it is stored.

```{r manifest-digest}
semantic <- fds_manifest_digest(manifest)
substr(semantic, 1, 16)

chunked <- fmri_frame(
  assays = list(bold = memory_source(signal, chunks = c(2L, 9L))),
  observations = observations,
  space = voxels,
  metadata = list(study = "toy")
)
identical(fds_manifest_digest(fds_frame_manifest(chunked)), semantic)
```

The digest is also unchanged when the *values* change, because values are not
semantics. Rename one observation, and it changes.

```{r manifest-digest-values}
doubled <- fmri_frame(
  assays = list(bold = signal * 2),
  observations = observations,
  space = voxels,
  metadata = list(study = "toy")
)
identical(fds_manifest_digest(fds_frame_manifest(doubled)), semantic)

renamed <- fmri_frame(
  assays = list(bold = signal),
  observations = transform(observations, .obs_id = sub("vol", "v", .obs_id)),
  space = voxels,
  metadata = list(study = "toy")
)
identical(fds_manifest_digest(fds_frame_manifest(renamed)), semantic)
```

Frames with ephemeral IDs have no manifest at all: `fds_frame_manifest()`
refuses them, so nothing random can reach disk.

## Identity domains

Identity is typed. `identity_descriptor()` returns a digest together with the
domain it belongs to and the canonicalization contract that produced it, so
two hexadecimal strings from different domains are never confused.

| Domain | Answers | Operation |
|---|---|---|
| semantic | Is this the same frame, ignoring storage? | `fds_manifest_digest()` |
| schema | Do these frames have the same column and space contracts? | `frame_schema_digest()` |
| space | Is this exactly the same feature space? | `space_digest()`, `same_space()` |
| source | Is this the same descriptor of the same revision of the same physical thing? | `source_fingerprint()` |
| provenance | Is this the same derivation history? | `provenance_digest()` |
| content | Do these arrays hold the same values? | `content_hash()`, passed in explicitly |

The domain is inferred from the object when it can be.

```{r identity}
identity_descriptor(frame)$domain
identity_descriptor(space(frame))$domain
identity_descriptor(assay(frame)$source)$domain
identity_descriptor(frame)$canonicalization$id
```

`explain()` reports the schema, semantic, and source identities of a frame
side by side, and none of them costs a read.

## Fingerprints are not content hashes

A source has two identities, computed separately, and the package never
substitutes one for the other.

`source_fingerprint()` is a cheap *revision* fingerprint of the descriptor:
its type, shape, dtype, chunk grid, selectors, and whatever physical evidence
the backend can observe without reading values, such as a file's size and
modification time. It is computed once at construction and cached. For an
in-memory source there is no file to observe, so the fingerprint includes a
per-object identity token. Two memory sources built from the same matrix are
two objects, and they have two fingerprints.

```{r fingerprint}
a <- memory_source(signal)
b <- memory_source(signal)
identical(source_fingerprint(a), source_fingerprint(b))
```

`content_hash()` is the one operation that identifies an array by its values.
It streams the source in bounded blocks and is proportional to the number of
values, so nothing calls it implicitly. It agrees across copies, storage
dtypes, chunk grids, and compositions of the same values.

```{r content-hash}
identical(content_hash(a), content_hash(b))

as_float32 <- memory_source(signal, dtype = "float32", chunks = c(1L, 18L))
identical(content_hash(a), content_hash(as_float32))

stacked <- row_bound_source(list(memory_source(signal[1:2, ]), memory_source(signal[3:6, ])))
identical(content_hash(a), content_hash(stacked))
```

A fingerprint says *same descriptor, same revision*. A content hash says *same
numbers*. Neither implies the other. If you want equal-valued memory sources
to share a fingerprint, opt in with `identity = "content"` and pay the hashing
cost once, visibly, at construction. A `revision` label replaces the
per-object token, so two memory sources built independently under the same
revision share a fingerprint without any hashing: the caller is asserting
that they are the same source in the same revision.

```{r fingerprint-opt-in}
identical(
  source_fingerprint(memory_source(signal, identity = "content")),
  source_fingerprint(memory_source(signal, identity = "content"))
)
```

A content hash becomes a typed identity by being handed to
`identity_descriptor()` explicitly. That is the only way the content domain is
ever populated.

```{r identity-content}
receipt <- content_hash(assay(frame)$source)
identity_descriptor(assay(frame)$source, domain = "content", content_digest = receipt)$digest ==
  receipt
```

File-backed sources use their fingerprint evidence to notice change: every
open and read re-observes the file and raises
`fmridataset_error_source_stale` when it no longer matches, distinct from the
`fmridataset_error_backend_io` a genuine I/O failure raises. The policy is
recorded in `inst/architecture/ADR-009-source-fingerprints-and-content-hashes.md`.

## HDF5 round trip

`write_frame()` persists a frame as an HDF5 file through the `fmristore`
package, which owns the certified layout and the atomic commit.
`open_frame()` reads it back. Reopened assays are lazy HDF5 sources; opening a
frame reads its manifest, not its values. Neither function computes a content
hash.

```{r write, eval = has_fmristore}
path <- file.path(tempdir(), "toy-frame.h5")
committed <- write_frame(frame, path)
reopened <- open_frame(committed)
reopened
class(assay(reopened)$source)[1]
```

The reopened frame has the same semantic identity, the same space, the same
IDs, the same run structure, and, when you ask for it, the same values.

```{r round-trip, eval = has_fmristore}
identical(fds_manifest_digest(fds_frame_manifest(reopened)), semantic)
same_space(space(reopened), space(frame))$same
identical(observation_ids(reopened), observation_ids(frame))
temporal_schema(reopened)$run_lengths
identical(content_hash(assay(reopened)$source), receipt)
```

What it does not share is a source fingerprint, because an HDF5 dataset and an
in-memory matrix are different physical things.

```{r round-trip-fingerprint, eval = has_fmristore}
identical(
  source_fingerprint(assay(reopened)$source),
  source_fingerprint(assay(frame)$source)
)
```

Views over the reopened frame behave exactly as they do in memory.

```{r round-trip-view, eval = has_fmristore}
collect_assay(reopened[c("vol-06", "vol-01"), c("voxel-2", "voxel-1")])
```

```{r write-cleanup, include = FALSE, eval = has_fmristore}
unlink(committed)
```

```{r no-fmristore, echo = FALSE, eval = !has_fmristore, results = "asis"}
cat("*The HDF5 examples were not evaluated because `fmristore` with frame support is not installed.*\n")
```

## Import from BIDS

`read_bids_bold()` opens one subject's preprocessed BOLD runs from an fMRIPrep
derivative tree as a single lazy frame. It uses the `bidser` package for
discovery. Construction reads the BOLD headers and the run masks, resolves one
common volume space, and mints deterministic observation IDs from the scan
path and volume index. It does not read BOLD values.

```{r bids-call, eval = FALSE}
bold <- read_bids_bold(
  "/data/my-study",
  subject = "01",
  task = "memory",
  space = "MNI152NLin2009cAsym"
)
```

To show the result without a real dataset, the chunk below writes a minimal
fMRIPrep-style tree with two three-volume runs into a temporary directory. The
helper is the same fixture the package's tests use.

```{r bids-fixture, eval = has_bidser}
make_fixture <- function(root = tempfile("bids-")) {
  func_dir <- file.path(root, "derivatives", "fmriprep", "sub-01", "func")
  dir.create(func_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(file.path(root, "sub-01", "func"), recursive = TRUE, showWarnings = FALSE)
  writeLines('{"Name":"toy","BIDSVersion":"1.10.0"}', file.path(root, "dataset_description.json"))
  writeLines(
    '{"Name":"fMRIPrep","BIDSVersion":"1.10.0","DatasetType":"derivative","GeneratedBy":[{"Name":"fMRIPrep"}]}',
    file.path(root, "derivatives", "fmriprep", "dataset_description.json")
  )
  write.table(
    data.frame(participant_id = "sub-01"), file.path(root, "participants.tsv"),
    sep = "\t", row.names = FALSE, quote = FALSE
  )
  grid <- c(2L, 2L, 2L)
  masks <- list(
    array(c(TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, FALSE, FALSE), grid),
    array(c(FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, FALSE), grid)
  )
  for (run in 1:2) {
    stem <- sprintf("sub-01_task-memory_run-%02d_space-MNI152NLin6Asym", run)
    bold <- array(seq_len(prod(grid) * 3L) + 100L * (run - 1L), c(grid, 3L))
    neuroim2::write_vec(
      neuroim2::NeuroVec(bold, neuroim2::NeuroSpace(c(grid, 3L))),
      file.path(func_dir, paste0(stem, "_desc-preproc_bold.nii"))
    )
    writeLines('{"RepetitionTime":2}', file.path(func_dir, paste0(stem, "_desc-preproc_bold.json")))
    neuroim2::write_vol(
      neuroim2::LogicalNeuroVol(masks[[run]], neuroim2::NeuroSpace(grid)),
      file.path(func_dir, paste0(stem, "_desc-brain_mask.nii"))
    )
    write.table(
      data.frame(onset = 2 * (run - 1), duration = 1, trial_type = c("old", "new")[run]),
      file.path(root, "sub-01", "func", sprintf("sub-01_task-memory_run-%02d_events.tsv", run)),
      sep = "\t", row.names = FALSE, quote = FALSE
    )
  }
  root
}
root <- make_fixture()
```

```{r bids-read, eval = has_bidser}
bold <- read_bids_bold(root, subject = "01", task = "memory", space = "MNI152NLin6Asym")
bold
```

The two run masks differ, and by default the frame uses their intersection:
three voxels. No resampling and no alignment happen implicitly; ambiguous
spaces, masks, or multi-echo selections are errors that ask for a choice.

```{r bids-space, eval = has_bidser}
feature_ids(bold)
```

The observation table carries the BIDS entities, the within-run volume index,
the run time, and the `TR`, so the temporal contract holds without further
work. A `run` entity keyed on `scan_id` and a `subject` entity are attached,
and `scan_id` is what the temporal schema uses as the run column, because a
BIDS `run` label is only unique within a session.

```{r bids-observations, eval = has_bidser}
observations(bold)[, c("run_id", "volume_index", "run_time", "TR")]
entities(bold)
temporal_schema(bold)$columns$run
unname(temporal_schema(bold)$run_lengths)
```

Observation IDs are durable and deterministic: the same tree imported twice
yields the same IDs.

```{r bids-ids, eval = has_bidser}
ids_are_durable(bold)
basename(observation_ids(bold)[1:2])
```

Reading is lazy and goes through the NIfTI source. A run filter reads only
that run's volumes, and because the frame keeps the complete feature domain of
its volume space, spatial maps can take the native read path.

```{r bids-read-values, eval = has_bidser}
collect_assay(filter_obs(bold, run_id == "run-2"))
execution_path(bold, "spatial")
```

Matching `events.tsv` files are attached as a keyed event table.

```{r bids-events, eval = has_bidser}
table_data(bold$tables$events)[, c("onset", "duration", "trial_type")]
```

```{r bids-cleanup, include = FALSE, eval = has_bidser}
unlink(root, recursive = TRUE)
```

```{r no-bidser, echo = FALSE, eval = !has_bidser, results = "asis"}
cat("*The BIDS examples were not evaluated because `bidser` (>= 0.5.0) is not installed.*\n")
```

## Choosing the right identity

Use the semantic digest to ask whether two frames mean the same thing, and the
schema digest to ask whether they could be bound or collected together. Use
`same_space()` before any operation that assumes columns line up. Key caches
and plans on the source fingerprint. Reach for `content_hash()` only when the
question is about the numbers, accept its cost knowingly, and record the
receipt where you need it, because the package will not do it for you.
