---
title: "Study-Level Analysis: From Single Subjects to Group Studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Study-Level Analysis: From Single Subjects to Group Studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  warning = FALSE,
  message = FALSE
)
library(fmridataset)
```

```{r make-subjects, include=FALSE}
set.seed(42)
make_subject <- function(id, n_voxels = 500) {
  set.seed(as.integer(chartr("0123456789", "0123456789", gsub("\\D", "", id))))
  matrix_dataset(
    datamat   = matrix(rnorm(200 * n_voxels), nrow = 200, ncol = n_voxels),
    TR        = 2.0,
    run_length = c(100, 100)
  )
}
subject_ids   <- c("sub-01", "sub-02", "sub-03", "sub-04", "sub-05")
subject_list  <- setNames(lapply(subject_ids, make_subject), subject_ids)
```

Multi-subject fMRI studies require a unified interface that scales from a
single participant to a full cohort without rewriting analysis code.
`fmridataset` provides `fmri_study_dataset()` for exactly this purpose:
wrap per-subject datasets into one object, then apply the same methods you
already use on single subjects.

## Create per-subject datasets

Each subject is an ordinary `matrix_dataset` (or any other backend — NIfTI,
HDF5, Zarr).  Create them once and collect them in a named list.

```{r per-subject}
ds_01 <- matrix_dataset(
  datamat    = matrix(rnorm(200 * 500), nrow = 200, ncol = 500),
  TR         = 2.0,
  run_length = c(100, 100)
)
ds_02 <- matrix_dataset(
  datamat    = matrix(rnorm(200 * 500), nrow = 200, ncol = 500),
  TR         = 2.0,
  run_length = c(100, 100)
)
```

## Compose into fmri_study_dataset()

Pass a named list of datasets and matching subject identifiers.

```{r compose-study}
study <- fmri_study_dataset(
  datasets    = subject_list,
  subject_ids = subject_ids
)
print(study)
```

The study object stores a lazy reference to each subject's data; nothing is
loaded into memory until you explicitly request it.

## Access data

Retrieve the full concatenated matrix, or pull one subject at a time.

```{r access-all}
all_data <- get_data_matrix(study)
dim(all_data)   # (total timepoints) x (voxels)
```

```{r access-one}
s01 <- get_data_matrix(study, subject_id = "sub-01")
dim(s01)        # timepoints for sub-01 x voxels
```

Subject-specific access is the most memory-efficient pattern for large
cohorts: load, compute, discard, repeat.

## Chunked iteration

`data_chunks()` returns an iterator that yields successive voxel blocks.
Use it when the full data matrix does not fit in memory.

```{r data-chunks}
chunks <- data_chunks(study, nchunks = 4)
repeat {
  ch <- tryCatch(iterators::nextElem(chunks), error = function(e) NULL)
  if (is.null(ch)) break
  cat("chunk", ch$chunk_num, ":", ncol(ch$data), "voxels\n")
}
```

Each chunk element exposes `$data` (timepoints x voxels) and
`$voxel_ind` (the column indices this chunk corresponds to).

## Group operations with fmri_group()

`fmri_group()` wraps a data-frame with one row per subject and a list
column of datasets.  It is the entry point for group-level modelling and
participant-level metadata.

```{r fmri-group}
subject_table <- data.frame(
  sub     = subject_ids,
  age     = c(24, 28, 31, 25, 29),
  dataset = I(lapply(subject_ids, function(id) list(subject_list[[id]]))),
  stringsAsFactors = FALSE
)

grp <- fmri_group(subject_table, id = "sub", dataset_col = "dataset")
print(grp)
```

With `grp` in hand you can attach demographic covariates, filter subjects
with `filter_subjects()`, and pass the group object to modelling functions
in downstream packages such as **fmrireg**.

## Next steps

- `vignette("bids-h5-archive")` — build a compressed, chunked HDF5 archive
  for an entire BIDS study, then read it back through the same interface
  shown here.
- `?fmri_study_dataset` — full argument reference and constructor options.
- `?data_chunks` — chunk size strategies, run-wise vs. voxel-wise iteration.
- `?fmri_group` — group-level metadata and subject filtering.

```{r session-info, include=FALSE}
sessionInfo()
```
